prolfqua - An R-package for Proteomics Differential Expression Analysis.

### 2022-05-04 inst/MyArticle/paper.Rmd
rename intensity to abudance. move to knitr tables.

---
title: prolfqua - An R-package for Proteomics Differential Expression Analysis.
author: 
  - Paolo Nanni$^{\star}$
  - Jonas Grossmann$^{\star}$
  - Maria d'Errico$^{\star}$
  - Ralph Schlapbach$^{\star}$
  - Christian Panse$^{\star}$
  - Witold E. Wolski $^{\star}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - Swiss Federal Institute of Technology in Zurich \| University of Zurich; Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. Correspondence$^\star$\texttt{wew@fgcz.ethz.ch}]
output: 
  bookdown::pdf_document2:
    toc: true
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
    Protein mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. Nevertheless, there is still a need for a flexible and easy-to-use application programmer interface in R that transparently supports a variety of well principled statistical inference procedures.
    The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package integrates essential steps of the mass spectrometry-based label-free quantification data analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The package can model simple experimental designs with a single explanatory variable and complex experiments with multiple factors using various statistical models. It also can be used to develop applications that produce interactive data analysis reports in PDF or HTML format. 
     Furthermore, the package can be used to benchmark data acquisition, data preprocessing, or data modeling methods using gold standard datasets. The R package is available through https://github.com/fgcz/prolfqua.

@@

We focused on comparing models to estimate fold changes, variances and to obtain false discovery rates.

We assessed the performance of different modeling approaches implemented in prolfqua, `r BiocStyle::Biocpkg("MSstats")` and `r BiocStyle::Biocpkg("proDA")`, using a benchmark dataset whose ground truth is known. However, the dataset used has several limitations: first, the design is a parallel-group design, i.e., we could not benchmark how well interactions are modelled; secondly, essential sources of variation present in every experiment, such as the biochemical and biological, are not measured. The dataset captures only the variance from the chromatography, electro-spray, and mass spectrometric measurement method. Thus, while we can extrapolate some of the results obtained to more realistic datasets, we need to be careful not to over-interpret the results.

We can conclude that if we want to order the proteins according to the likelihood of being differentially regulated, for instance, to perform gene set enrichment analysis, the $t$-statistic is better suited than the fold-change estimate. Modeling the degrees of freedom when computing the $p$-values might improve the inference further. However, the observed change is minuscule.

For the fixed effect linear model, the empirical Bayes variance shrinkage, as suggested by [@Smyth2004linear], consistently improves compared with the unmoderated estimates (data not shown). However, there is no such effect for the mixed effect model, most likely because the degrees of freedom are erroneously estimated for a large proportion of models.

@@

The probabilistic dropout analysis implemented in the `r BiocStyle::Biocpkg("proDA")` produces inferences comparable to other methods (Figure \@ref(fig:FDRfdp) for the largest number of proteins (Figure \@ref(fig:FDRfdp) left panel). Moreover, the estimated fold changes are unbiased and show high diagnostic accuracy. Furthermore, the scaled $p$-values and $t$-statistics are on par with those obtained by the linear model with variance moderation. Therefore, we plan to integrate the `r BiocStyle::Biocpkg("proDA")` package as an additional modeling option into `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)).

Leveraging on these computer experiments, we can provide the following advice. First, aggregate the peptide intensities to protein intensities using a robust regression method. Second, use linear models as long as they suffice to model your experimental design. Use mixed effect models only if you have to, e.g., repeated measurements. If you use fixed-effect linear models, apply variance moderation since this improves the $t$-statistics and $p$-value estimates. If you want to sort your protein lists to perform gene set enrichment analysis, use the $t$-statistic instead of the fold change. You can select proteins to study them further based on the FDR, although expect a $bias$.

`r BiocStyle::Githubpkg("fgcz/prolfqua")` allows for considerable flexibility to model your experiments. Various types of models are available (see Figure \@ref(fig:ContrastUML) and Table \@ref(tab:prolfquaModels)), and the contrast specification is consistent and intuitive for all models. The benchmarking results are comparable with existing methods. The modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")`, allows for customizing the analysis or adding new features, e.g., generalized linear models (_glm_'s) for missing data modeling or robust linear models (_rlm_'s), quickly.

@@

@@

# Conclusion (tbd)

@@

x <- dplyr::inner_join(imp, proda, by = c("contrast","protein_Id", "TP", "species"), suffix = c(".imp",".proda"))
ggplot2::ggplot(x,ggplot2::aes( x = diff.imp, y = diff.proda) ) + ggplot2::geom_point() + ggplot2::facet_wrap(~contrast + species) + ggplot2::geom_hline(yintercept = 0, color="red") + ggplot2::geom_vline(xintercept = 0, color="red")
```

>>>>

---
title: prolfqua - An R-package for Proteomics Differential Expression Analysis.
author: 
  - Paolo Nanni$^{\star}$
  - Jonas Grossmann$^{\star}$
  - Maria d'Errico$^{\star}$
  - Ralph Schlapbach$^{\star}$
  - Christian Panse$^{\star}$
  - Witold E. Wolski $^{\star}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - Swiss Federal Institute of Technology in Zurich \| University of Zurich; Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. Correspondence$^\star$\texttt{wew@fgcz.ethz.ch}]
output: 
  bookdown::pdf_document2:
    toc: true
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
    Protein mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. Nevertheless, there is still a need for a flexible and easy-to-use application programmer interface in R that transparently supports a variety of well principled statistical inference procedures.
    The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package integrates essential steps of the mass spectrometry-based label-free quantification data analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The package can model simple experimental designs with a single explanatory variable and complex experiments with multiple factors using various statistical models. It also can be used to develop applications that produce interactive data analysis reports in PDF or HTML format. 
     Furthermore, the package can be used to benchmark data acquisition, data preprocessing, or data modeling methods using gold standard datasets. Using the benchmark functionality, we show that the methods implemented allow performing sensitive and specific differential expression analysis. 
     The R package is available through https://github.com/fgcz/prolfqua.

@@

We assessed the performance of different modeling approaches implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")`, `r BiocStyle::Biocpkg("MSstats")` and `r BiocStyle::Biocpkg("proDA")`, using a benchmark dataset whose ground truth is known. However, since the for each dilution only technical replicates are available, essential sources of variation present in every experiment, such as the biochemical and biological, are not measured. The dataset captures only the variance from the chromatography, electro-spray, and mass spectrometric measurement method. Thus, while we can extrapolate some of the results obtained to more realistic datasets, we need to be careful not to over-interpret the results. Specifically, in more realistic experiments the observed variances will be higher and sensitivities and specificity, will be worse given the same sample sizes. Furthermore, the proportion of missing observations in real world datasets might be higher.

However, we can conclude that if we want to order the proteins according to the likelihood of being differentially regulated, to perform gene set enrichment analysis [@subramanian2005gene], the $t$-statistic is better suited than the fold-change estimate. Modeling the degrees of freedom when computing the $p$-values might further improve the inference, However, the observed change is minuscule (see Figure \@ref(fig:)). Furthermore, for the fixed effect linear model, the empirical Bayes variance shrinkage, as suggested by [@Smyth2004linear], consistently improves the ranking of proteins compared with the unmoderated estimates (data not shown). However, there is no such effect for the mixed effect model, most likely because the degrees of freedom are erroneously estimated for a large proportion of models.

@@

The probabilistic dropout analysis implemented in the `r BiocStyle::Biocpkg("proDA")` produces inferences comparable to other methods (Figure \@ref(fig:FDRfdp) for the largest number of proteins (Figure \@ref(fig:FDRfdp) left panel). Moreover, the estimated fold changes are unbiased and show high diagnostic accuracy. Furthermore, the scaled $p$-values or the $t$-statistics are on par with those obtained by the linear model with variance moderation. Therefore, we plan to integrate the `r BiocStyle::Biocpkg("proDA")` package as an additional modeling option into `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)).

@@

# Conclusion

`r BiocStyle::Githubpkg("fgcz/prolfqua")` allows for considerable flexibility to model your experiments. Various types of models are available (see Figure \@ref(fig:ContrastUML) and Table \@ref(tab:prolfquaModels)), and the contrast specification is consistent and intuitive for all models. The modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")`, allows for adding new features, e.g., generalized linear models (_glm_'s) for missing data modeling or robust linear models (_rlm_'s), in the future.

Leveraging on these computer experiments, we can provide the following advice. First, aggregate the peptide intensities to protein intensities using a robust regression method. Second, use linear models as long as they suffice to model your experimental design. Use mixed effect models only if you have to, e.g., repeated measurements and if the sample sizes are large. If you use fixed-effect linear models, apply variance moderation since this improves the $t$-statistics and $p$-value estimates. If you want to sort your protein lists to perform gene set enrichment analysis, use the $t$-statistic instead of the fold change. You can select proteins to study them further based on the FDR, although expect a $bias$.

@@

@@

x <- dplyr::inner_join(imp, proda, by = c("contrast","protein_Id", "TP", "species"), suffix = c(".imp",".proda"))
ggplot2::ggplot(x,ggplot2::aes( x = diff.imp, y = diff.proda) ) + ggplot2::geom_point() + ggplot2::facet_wrap(~contrast + species) + ggplot2::geom_hline(yintercept = 0, color="red") + ggplot2::geom_vline(xintercept = 0, color="red")
